\section{Quantum operations}
\label{sec:quantum_operations}
\subsection{Single-qubit gates}
To operate on one or more qubits, a unitary operation is applied to the state, where the unitarity is needed for states to remain normalised.
As they are unitary, any purity is preserved.
With the finite number of qubits, these operations can be expressed as matrices.
These operations are often thought of as gates, paralleling the classical gates in digital logic.


The most basic gates are the Pauli gates, which are applications of the Pauli matrices from \cref{eq:pauli} and are as gates simply denoted as $X$, $Y$, and $Z$ respectively.
These gates can be seen as half turns around the $x$-, $y$- and $z$-axes, respectively, of the Bloch sphere.
The $X$-gate is also known as the NOT gate, as it mirrors the classical NOT gate by mapping $\ket{0}$ to $\ket{1}$ and vice versa, though it of course is more general, being also applicable to superposition states.

The Hadamard gate
\begin{equation}
    H = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 & 1 \\ 1 & -1 \end{pmatrix}
\end{equation}
is a rotation around the $x$-axis by $\pi/2$.
It is a very important gate in quantum computing, as it is used to create superpositions of the computational basis states.


The $R_X$-, $R_Y$- and $R_Z$-gates are rotations around the $x$-, $y$- and $z$-axes, respectively, by an arbitrary angle $\theta$:
\begin{align*}
    R_X(\theta) & = \begin{pmatrix} \cos\left(\frac{\theta}{2}\right) & -i \sin\left(\frac{\theta}{2}\right) \\ -i \sin\left(\frac{\theta}{2}\right) & \cos\left(\frac{\theta}{2}\right) \end{pmatrix}, \\
    R_Y(\theta) & = \begin{pmatrix} \cos\left(\frac{\theta}{2}\right) & -\sin\left(\frac{\theta}{2}\right) \\ \sin\left(\frac{\theta}{2}\right) & \cos\left(\frac{\theta}{2}\right) \end{pmatrix},      \\
    R_Z(\theta) & = \begin{pmatrix} e^{-i\frac{\theta}{2}} & 0 \\ 0 & e^{i\frac{\theta}{2}} \end{pmatrix}.
\end{align*}
These parametrised gates will be useful in \cref{sec:vqa}.

\subsection{Multi-qubit gates}
The most important multi-qubit gate is the controlled $X$-gate, also known as the CNOT, which is a controlled version of the $X$-gate.
Being controlled means that it only acts on the second qubit if the first qubit is in the state $\ket{1}$.
Of course, the first qubit may be in a superposition, and the CNOT this way allows for the creation of entanglement between the two qubits.
If the first qubit has probability amplitude $\alpha$ of being in the state $\ket{1}$, the second qubit will have probability amplitude $\alpha$ of being flipped.
The CNOT gate can be expressed in matrix form as
\begin{equation}
    \text{CNOT} = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 0 & 1 & 0 \end{pmatrix}.
\end{equation}

In theory, any unitary single-qubit operation can be controlled.
Another interesting two-qubit gate is the controlled $Z$-gate, CZ, expressible as the matrix
\begin{equation}
    \text{CZ} = \begin{pmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 1 & 0 \\ 0 & 0 & 0 & -1 \end{pmatrix}.
\end{equation}
Because it only alters the amplitude of $\ket{11}$, it does not actually matter which qubit is the control and which is the target.



\subsection{Observables and measurements}
For an output to be obtained from a quantum computer, a measurement must be performed.
This is typically done at the end of all operations and of all qubits, thus yielding a single output of zeroes and ones, a bit-string.
It is important to note that the measurement is not a deterministic process, but rather a probabilistic one.
Often, the underlying probabilities are what is of interest.
Therefore, many measurements are performed.
Usually, these results are averaged to obtain an estimate, but more complicated post-processing methods are also possible.
For instance, neural networks have shown useful properties in regard of reducing variance in the estimates, though at the cost of some bias \cite{torlai2020}.

The $Z$-basis is the canonical basis for measurements, but any other basis can be used, at least in theory.
Often, only canonical basis measurements are implemented in the hardware.
Using another basis can be done by properly preparing the state before the measurement.

Measurements may be done in the middle of operations and be used to control the operations.
If the qubits are entangled, measuring one will affect the measurement probabilities of others.
Using such intermediate measurements is a way of introducing non-linearities in the otherwise unitary nature of the quantum world.


\subsection{Quantum circuits}
The operations on qubits are often described using quantum circuits, which are a graphical representation of the operations on the qubits, the quantum algorithms.
They are read from left to right.
It is standard procedure to assume all qubits start in the state $\ket{0}$.
Gates are generally written as boxes with the name of the gate inside.

A simple example is the circuit
\begin{equation}
    \begin{quantikz}
        \lstick{$\ket{0}$} & \gate{H} & \qw \\
        \lstick{$\ket{0}$} & \gate{H} & \qw
    \end{quantikz},
\end{equation}
which prepares the state $\frac{1}{2}(\ket{00} + \ket{01} + \ket{10} + \ket{11})$.
This is a pure state with no entanglement, and so the measurement probabilities of the two qubits are independent.

Slightly more interesting is the circuit
\begin{equation}
    \begin{quantikz}
        \lstick{$\ket{0}$} & \gate{H} & \ctrl{1} & \qw \\
        \lstick{$\ket{0}$} & \qw & \targ{} & \qw
    \end{quantikz}
\end{equation}
in which the first qubit is put into a superposition using the Hadamard gate before a CNOT gate is applied to the second, controlled on the first.
This creates the state $\frac{1}{\sqrt{2}}\left(\ket{00} + \ket{11}\right)$.
The measurement probabilities of the two qubits are now correlated; if the first qubit is measured to be $\ket{1}$, the second will always be $\ket{1}$ and vice versa.
The probability of measuring the qubits to be different is nil.

To create a mixed state, an intermediate measurement can be used to control a gate.
For instance, the circuit
\begin{equation}
    \begin{quantikz}
        \lstick{$\ket{0}$} & \gate{H} & \meter{} & \cwbend{1} \\
        \lstick{$\ket{0}$} & \qw & \qw & \gate{X} & \qw
    \end{quantikz}
\end{equation}
places the second qubit in the mixed state $\frac{1}{2}(\ketbra{0}{0} + \ketbra{1}{1})$.
If it were immediately to be measured, it would have a 50\% chance of being $\ket{0}$ and a 50\% chance of being $\ket{1}$.
The uncertainty is only classical, and it could therefore not be used to create entanglement or for any other quantum spookiness.

\subsection{Quantum supremacy}
\begin{sloppypar}
    Exponential speed-ups do not come for free.
    Quantum computers do only solve certain problems more efficiently than classical computers, and finding the algorithms to do so is no easy task.
    \sloppy Shor's algorithm has time complexity $O((\log N)^2 (\log \log N) (\log \log \log N))$ while the most efficient known classical algorithm has time complexity $O(\exp{1.9(\log N)^{1/3} \text{} (\log \log N)^{2/3}})$.
    To solve linear system algorithms, the HHL algorithm has time complexity $O(\log(N)\kappa^2)$ where $\kappa$ is the condition number, which is an exponential speed-up over the fastest known classical algorithm has time complexity $O(N \kappa)$.
    Still, discovering these algorithms is not trivial.
\end{sloppypar}

Polynomial speed-ups are also found.
For example, the Grover algorithm which is used to search for an element in an unsorted list has time complexity $O(\sqrt{N})$.
Classically, this can not be done in less than $O(N)$ time.
This algorithm, or the more general amplitude amplification on which it builds, solves a very general problem and is often used a subroutine to achieve quadratic speed-ups in other algorithms.
Being only a quadratic speed-up, it is not as impressive as the exponential speed-ups,
and achieving quantum supremacy in that manner would require larger quantum computers than if the speed-up were exponential.

It is proven that the class of problems quantum computers can solve in polynomial time contains P.
This follows from the fact than quantum computers can do any classical algorithm by also solving problems like discrete logarithm factorisation in polynomial time.
However, it is not known if quantum computers can solve NP-hard problems in efficiently.