\pdfbookmark[0]{Preface}{Preface}
\section*{Preface}
This report concludes the course \textit{TMA4500 Industrial mathematics, specialisation project} whose purpose is two-fold.
First, it is to acquaint the author-student to a specific topic, which is this case is quantum machine learning or more generally quantum computing.
Secondly, it is to make the student familiar with the process of writing a semester-long thesis under the supervision of an expert, and so make them ready for the master's thesis next semester.

The initial aim of the project was to learn about variational quantum algorithms and the application thereof to machine learning.
This lead to the investigation of quantum neural networks, which are mostly just interpretation of variational quantum algorithms through the lens of standard, classical neural network design.
With the limited scope of this pre-master thesis, it was decided only to implement and test some simple models based on recent research papers.

I would like to thank my de-facto supervisor Franz Fuchs for not only taking time off his important work to supervise me, but also for showing enthusiasm and inviting me into this brave new world of quantum computing when I did not really know what I wanted to do.
I also want to thank my co-supervisor, Alexander Stasik, for his help and guidance in the process of writing this thesis.
Finally, I would like to thank my official supervisor, Gunnar Taraldsen, for allowing me to work with something I find truly exciting and for still providing feedback while I worked with something out of the ordinary.

All code used in this report will be made available on GitHub via \url{https://github.com/boyesjo/tma4500}.



\cleardoublepage